Hypercoercions and a Framework for Equivalence of Cast Calculi

### 2019-10-21 paper.tex
edits to section 2, start of adding text to sec 4

\newcommand{\typingHC}[3]{#1 : #2 \Longrightarrow #3}
\newcommand{\hcvOOinj}[2]{\mathtt{inj} \; #2}
\newcommand{\hcvOOfun}[5]{\mathtt{fun} \; #1 \; #2 \; #3 \; #4 \; #5}
\newcommand{\hcvOOtt}[0]{\mathtt{unit}}
\newcommand{\hcvOOcons}[4]{\mathtt{cons}\;#1\;#2\;#3\;#4}
\newcommand{\hcvOOinl}[2]{\mathtt{inl}\;#1\;#2}
\newcommand{\hcvOOinr}[2]{\mathtt{inr}\;#1\;#2}
\newcommand{\hckOOmt}[0]{\mathtt{stop}}
\newcommand{\hckOOconsI}[3]{\mathtt{cons_1}\;#1\;#2\;#3}
\newcommand{\hckOOappII}[2]{\mathtt{app_2}\;#1\;#2}
\newcommand{\sidecond}[1]{\text{if}\;#1}
% Lazy D cast calculus on space-inefficient CEK
\newcommand{\judgeCreduce}[2]{#1 \longmapsto_{\mathcal{C}} #2}
\newcommand{\judgeCreduceTrans}[2]{#1 \longmapsto_{\mathcal{C}}^{*} #2}
\newcommand{\judgeCeval}[2]{eval_{\mathcal{C}}(#1) = #2}
\newcommand{\redrule}[3]{#1 & \longmapsto_\mathcal{C} & #2 & #3\\}
\newcommand{\hiredrule}[3]{\highlight{#1} & \highlight{\longmapsto_\mathcal{C}} & \highlight{#2} & \highlight{#3} \\}
% blame calculus on space-efficient CEK
\newcommand{\judgeSreduce}[3]{#2 \longmapsto_{\mathcal{S}(#1)} #3}
\newcommand{\judgeSreduceTrans}[3]{#2 \longmapsto_{\mathcal{S}(#1)}^{*} #3}
\newcommand{\judgeSeval}[3]{eval_{\mathcal{S}(#1)}(#2) = #3}
\newcommand{\redruleS}[3]{#1 & \longmapsto_{\mathcal{S}(C)} & #2 & #3\\}
\newcommand{\hiredruleS}[3]{\highlight{#1} & 
\highlight{\longmapsto_{\mathcal{S}(C)}} & \highlight{#2} & \highlight{#3} \\}
% Normal Coercion
\newcommand{\ncProj}[2]{#1?^{#2}}
\newcommand{\ncInj}[1]{#1!}
\newcommand{\ncId}[0]{\iota}
\newcommand{\ncSeq}[2]{#1;#2}
\newcommand{\ncFail}[1]{\bot^{#1}}
\newcommand{\ncFun}[2]{\POOfun{#1}{#2}}
\newcommand{\ncProd}[2]{\POOprod{#1}{#2}}
\newcommand{\ncSum}[2]{\POOsum{#1}{#2}}
% Hypercoercion
\newcommand{\hyperCoercionI}[0]{\mathtt{id\star}}
\newcommand{\hyperCoercionC}[3]{#1 \overset{#2}{\curvearrowright} #3}
% machine state simulations
\newcommand{\eqvS}[4]{#3 \approx_{\mathcal{S}\mathcal{S}} #4}
\newcommand{\eqvSD}[3]{#2 \approx_{\mathcal{SD}} #3}
% to-dos
\newcommand{\todoKC}[1]{\todo[inline]{KC needs to #1}}
\newcommand{\todoKCFixed}[0]{\todo[inline]{Fixed. -KC}}
% abbrev
\newcommand{\castCalculus}[0]{$\lambda_{\rightarrow}^{\langle\cdot\rangle}$}
% names
\newcommand{\ineffCEK}{$ \mathcal{C} $}
\newcommand{\ineffCEKD}{$ \mathcal{D} $}
\newcommand{\ineffCEKUD}{$ \mathcal{UD} $}
\newcommand{\judgeDreduce}[2]{#1 \longmapsto_{\mathcal{D}} #2}
\newcommand{\judgeDreduceTrans}[2]{#1 \longmapsto_{\mathcal{D}}^{*} #2}
\newcommand{\judgeDeval}[2]{eval_{\mathcal{D}}(#1) = #2}
\newcommand{\effCEK}[1]{$ \mathcal{S}(#1) $}
\newcommand{\evalEqv}[2]{$ eval_{\text{#1}} = eval_{\text{#2}} $}
\newcommand{\continue}[2]{continue(#2,#1)}
\newcommand{\highlight}[1]{{\color{red} #1}}

@@

\todo[inline]{There are missing transition rules for fst and snd. -JS
\\---\\
Fixed. -KC}

@@

\subsection{The \lazyD{} cast calculus}

@@

\figref{fig:applyCast-D-C} defines the $ applyCast $ for \lazyD. 
We conjecture that with this $ applyCast $, the CEK machine in 
\figref{fig:machine-cekc} would agree with the \lazyD{} cast calculus in 
\citet{siek2009exploring}. We denote this \lazyD{} machine by \ineffCEKD{},
and the corresponding relations by $ \judgeDreduce{s}{s} $, $ 
\judgeDreduceTrans{s}{s} $, and $ \judgeDeval{e}{o} $

\subsection{The \lazyUD{} cast calculus [Jeremy]}

\todo[inline]{I need to update this subsection. -JS}

The dynamic semantics of the \lazyUD{} cast calculus, which we define here as a
CEK machine, is similar to that of the \lazyD{} cast calculus
(Fig. \ref{fig:machine-cekc}). The only difference is in the definition of
the $\mathit{applyCast}$ function, in which a cast whose source or
target is the unknown type $\star$ is always split into two casts that
go through an injectiable type, that is, a type in which all
sub-components are the unknown type, such as $\star \to \star$. The
definition of $\mathit{applyCast}$ for the \lazyUD{} cast calculus is given in
Fig.~\ref{fig:apply-Cast-UD}.

\begin{figure}
%
%  Syntax
%  \[
%  \begin{array}{rclr}
%    \stxrule{v}{values}{
%      \cdots \mid 
%      \vOOcast{v}{ \cOOcast{I}{l}{\star} } \mid
%      \vOOcast{v}{ \cOOcast{P}{l}{P} }
%    }
%  \end{array}
%  \]
%  
  \fbox{$\mathit{applyCast}(v,c) = r$}
  \[
  \begin{array}{rclr}
    \mathit{applyCast}(v, \cOOcast{\star}{l}{\star} ) &=& v \\
    \mathit{applyCast}(v, \cOOcast{P}{l}{\star}) &=&
        v \langle \cOOcast{P}{l}{I} \rangle
          \langle \cOOcast{I}{l}{\star} \rangle
        & \text{if } I \sim P, I \neq P \\  
    \mathit{applyCast}(v, \cOOcast{\star}{l}{P}) &=&          
        v \langle \cOOcast{\star}{l}{I} \rangle
          \langle \cOOcast{I}{l}{P} \rangle
        & \text{if } I \sim P, I \neq P \\  
  \mathit{applyCast}(v \langle \cOOcast{I}{l}{\star} \rangle , \cOOcast{\star}{l}{I}) &=& v \\
  \mathit{applyCast}(v \langle \cOOcast{I_1}{l}{\star} \rangle , \cOOcast{\star}{l}{I_2}) &=& \rOOfail{l} & \text{if } I_1 \neq I_2 \\
  \mathit{applyCast}(v, \cOOcast{P_1}{l}{P_2}) &=&
     v \langle \cOOcast{P_1}{l}{P_2} \rangle & \text{if } P_1 \smile P_2
  \end{array}
  \]

@@

@@

\section{a framework for proving correctness of cast representations}
\label{sec:framework}

This section presents our second contribution, a framework for proving 
correctness of cast representations. The framework is a theorem saying that for 
all cast representation $ C $, if $ C $ satisfies a collection of properties, 
then it would respect the Lazy D semantics of cast calculi. We are working on a 
similar theorem for \lazyUD{}.

@@

\todoKC{describe transition rules}
We list a fraction of reduction rules due to space limitation.
When values are constructed, their hypercoercion parts are filled with outputs 
of $ id $. For instance, when a function value is constructed, its first part 
and last part are initialized to $ id(T_1) $ and $ id(T_2) $ respectively.
When evaluating a cast expression, the current continuation is extended with a 
hypercoercion constructed by $ cast $. $ ext $ composes the new hypercoercion 
with the hypercoercion on the top of the continuation.
When evaluating a compound expression, the machine firstly construct the new 
pre-continuation, then turn it to a continuation by adding an identity 
hypercoercion at the top. For instance, when evaluating a \texttt{cons} 
expression, the machine firstly construct $ \hckOOconsI{e_2}{\rho}{\kappa} $, 
the new pre-continuation, then call $ cont $, which adds an identity cast to 
form a continuation. 
When a function call happens, the machine firstly cast the operand. If the 
casting succeeds, the machine then evaluate the body in the extended 
environment and the extended continuation. If the casting fails, the machine 
then halts with the blame label.

@@

\todo[inline]{Standardize theorem statement -KC}
\begin{lemma}[Strong Bi-simulation among $ \mathcal{S}(\cdot) $]
	\label{thm:CEKS-bisim}
	If 
	$ C_1 $ and $ C_2 $ are instances of Lazy D Cast ADT,
	$ s_1, s_1' \in S(C_1) $ and $ s_2,s_2' \in S(C_2) $,
	$ s_1 \approx s_2 $,
	$ \judgeSreduce{C_1}{s_1}{s_1'} $, and
	$ \judgeSreduce{C_2}{s_2}{s_2'} $, then 
	$ s_1' \approx s_2' $

>>>>

\newcommand{\typingHC}[3]{#1 : #2 \Longrightarrow #3}
\newcommand{\hcvOOinj}[2]{\mathtt{inj} \; #2}
\newcommand{\hcvOOfun}[5]{\mathtt{fun} \; #1 \; #2 \; #3 \; #4 \; #5}
\newcommand{\hcvOOtt}[0]{\mathtt{unit}}
\newcommand{\hcvOOcons}[4]{\mathtt{cons}\;#1\;#2\;#3\;#4}
\newcommand{\hcvOOinl}[2]{\mathtt{inl}\;#1\;#2}
\newcommand{\hcvOOinr}[2]{\mathtt{inr}\;#1\;#2}
\newcommand{\hckOOmt}[0]{\mathtt{stop}}
\newcommand{\hckOOconsI}[3]{\mathtt{cons_1}\;#1\;#2\;#3}
\newcommand{\hckOOappII}[2]{\mathtt{app_2}\;#1\;#2}
\newcommand{\sidecond}[1]{\text{if}\;#1}
% Lazy D cast calculus on space-inefficient CEK
\newcommand{\judgeCreduce}[2]{#1 \longmapsto_{\mathcal{C}} #2}
\newcommand{\judgeCreduceTrans}[2]{#1 \longmapsto_{\mathcal{C}}^{*} #2}
\newcommand{\judgeCeval}[2]{eval_{\mathcal{C}}(#1) = #2}
\newcommand{\redrule}[3]{#1 & \longmapsto_\mathcal{C} & #2 & #3\\}
\newcommand{\hiredrule}[3]{\highlight{#1} & \highlight{\longmapsto_\mathcal{C}} & \highlight{#2} & \highlight{#3} \\}
% blame calculus on space-efficient CEK
\newcommand{\judgeSreduce}[3]{#2 \longmapsto_{\mathcal{S}(#1)} #3}
\newcommand{\judgeSreduceTrans}[3]{#2 \longmapsto_{\mathcal{S}(#1)}^{*} #3}
\newcommand{\judgeSeval}[3]{eval_{\mathcal{S}(#1)}(#2) = #3}
\newcommand{\redruleS}[3]{#1 & \longmapsto_{\mathcal{S}(C)} & #2 & #3\\}
\newcommand{\hiredruleS}[3]{\highlight{#1} & 
\highlight{\longmapsto_{\mathcal{S}(C)}} & \highlight{#2} & \highlight{#3} \\}
% Normal Coercion
\newcommand{\ncProj}[2]{#1?^{#2}}
\newcommand{\ncInj}[1]{#1!}
\newcommand{\ncId}[0]{\iota}
\newcommand{\ncSeq}[2]{#1;#2}
\newcommand{\ncFail}[1]{\bot^{#1}}
\newcommand{\ncFun}[2]{\POOfun{#1}{#2}}
\newcommand{\ncProd}[2]{\POOprod{#1}{#2}}
\newcommand{\ncSum}[2]{\POOsum{#1}{#2}}
% Hypercoercion
\newcommand{\hyperCoercionI}[0]{\mathtt{id\star}}
\newcommand{\hyperCoercionC}[3]{#1 \overset{#2}{\curvearrowright} #3}
% machine state simulations
\newcommand{\eqvS}[4]{#3 \approx_{\mathcal{S}\mathcal{S}} #4}
\newcommand{\eqvSD}[3]{#2 \approx_{\mathcal{SD}} #3}
% to-dos
\newcommand{\todoKC}[1]{\todo[inline]{KC needs to #1}}
\newcommand{\todoKCFixed}[0]{\todo[inline]{Fixed. -KC}}
% abbrev
\newcommand{\castCalculus}[0]{$\lambda_{\rightarrow}^{\langle\cdot\rangle}$}
% names
\newcommand{\ineffCEK}{$ \mathcal{C} $}
\newcommand{\ineffCEKD}{$ \mathcal{D} $}
\newcommand{\ineffCEKUD}{$ \mathcal{UD} $}
\newcommand{\judgeDreduce}[2]{#1 \longmapsto_{\mathcal{D}} #2}
\newcommand{\judgeDreduceTrans}[2]{#1 \longmapsto_{\mathcal{D}}^{*} #2}
\newcommand{\judgeDeval}[2]{eval_{\mathcal{D}}(#1) = #2}
\newcommand{\judgeUDreduce}[2]{#1 \longmapsto_{\mathcal{UD}} #2}
\newcommand{\judgeUDreduceTrans}[2]{#1 \longmapsto_{\mathcal{UD}}^{*} #2}
\newcommand{\judgeUDeval}[2]{eval_{\mathcal{UD}}(#1) = #2}
\newcommand{\effCEK}[1]{$ \mathcal{S}(#1) $}
\newcommand{\evalEqv}[2]{$ eval_{\text{#1}} = eval_{\text{#2}} $}
\newcommand{\continue}[2]{continue(#2,#1)}
\newcommand{\highlight}[1]{{\color{red} #1}}

@@

@@

@@

\begin{definition}[\lazyD{} CEK Machine]
  The \lazyD{} CEK machine, written \ineffCEKD{}, is the CEK machine of
  \figref{fig:machine-cekc} using the $applyCast$ for \lazyD{} defined
  in \figref{fig:applyCast-D-C}.  We write the transition relations of
  this machine as $\judgeUDreduce{s}{s}$ and $\judgeUDreduceTrans{s}{s}$
  and write the evaluation function as $\judgeUDeval{e}{o}$.
\end{definition}

We conjecture that \ineffCEKD{} agrees with the \lazyD{} cast calculus
of \citet{siek2009exploring}.

Next we define the CEK Machine for \lazyUD{}. The only difference with
respect to \lazyD{} is in the definition of the $\mathit{applyCast}$
function, in which a cast whose source or target is the unknown type
$\star$ is always split into two casts that go through an injectiable
type, that is, a type in which all sub-components are the unknown
type, such as $\star \to \star$.

\begin{definition}[\lazyUD{} CEK Machine]
  The \lazyUD{} CEK machine, written \ineffCEKUD{}, is the CEK machine of
  \figref{fig:machine-cekc} using the $applyCast$ for \lazyUD{} defined
  in \figref{fig:apply-Cast-UD}.  We write the transition relations of
  this machine as $\judgeDreduce{s}{s}$ and $\judgeDreduceTrans{s}{s}$
  and write the evaluation function as $\judgeDeval{e}{o}$.
\end{definition}

\begin{figure}
  \fbox{$\mathit{applyCast}(v,c) = r$}
  \[
  \begin{array}{rclr}
    \mathit{applyCast}(v, \cOOcast{\star}{l}{\star} ) &=& v \\
    \mathit{applyCast}(v, \cOOcast{P}{l}{\star}) &=&
        v \langle \cOOcast{P}{l}{I} \rangle
          \langle \cOOcast{I}{l}{\star} \rangle
        & \text{if } I \sim P, I \neq P \\  
    \mathit{applyCast}(v, \cOOcast{\star}{l}{P}) &=&          
        v \langle \cOOcast{\star}{l}{I} \rangle
          \langle \cOOcast{I}{l}{P} \rangle
        & \text{if } I \sim P, I \neq P \\  
  \mathit{applyCast}(v \langle \cOOcast{I}{l}{\star} \rangle , \cOOcast{\star}{l}{I}) &=& v \\
  \mathit{applyCast}(v \langle \cOOcast{I_1}{l}{\star} \rangle , \cOOcast{\star}{l}{I_2}) &=& \rOOfail{l} & \text{if } I_1 \neq I_2 \\
  \mathit{applyCast}(v, \cOOcast{P_1}{l}{P_2}) &=&
     v \langle \cOOcast{P_1}{l}{P_2} \rangle & \text{if } P_1 \smile P_2
  \end{array}
  \]

@@

We conjecture that \ineffCEKUD{} agrees with the \lazyUD{} cast
calculus of \citet{siek2009exploring}.

@@

\section{A framework for proving correctness of cast representations}
\label{sec:framework}

This section presents our second contribution, a framework for proving
correctness of cast representations. The next section will apply this
framework to prove the correctness of the \lazyD{} hypercoercions.

The framework is a theorem saying that for all cast representation $ C
$, if $ C $ satisfies a collection of properties, then it would
respect the Lazy D semantics of cast calculi. We are working on a
similar theorem for \lazyUD{}.

@@

\todoKC{describe transition rules}
We list a fraction of reduction rules due to space limitation.
\todo[inline]{Don't omit reduction rules, instead put some in the appendix. -JS}
When values are constructed, their hypercoercion parts are filled with outputs 
of $ id $. For instance, when a function value is constructed, its first part 
and last part are initialized to $ id(T_1) $ and $ id(T_2) $ respectively.
When evaluating a cast expression, the current continuation is extended with a 
hypercoercion constructed by $ cast $. $ ext $ composes the new hypercoercion 
with the hypercoercion on the top of the continuation.
When evaluating a compound expression, the machine firstly construct the new 
pre-continuation, then turn it to a continuation by adding an identity 
hypercoercion at the top. For instance, when evaluating a \texttt{cons} 
expression, the machine firstly construct $ \hckOOconsI{e_2}{\rho}{\kappa} $, 
the new pre-continuation, then call $ cont $, which adds an identity cast to 
form a continuation. 
When a function call happens, the machine firstly cast the operand. If the 
casting succeeds, the machine then evaluate the body in the extended 
environment and the extended continuation. If the casting fails, the machine 
then halts with the blame label.

@@

\todo[inline]{Standardize theorem statement -KC}
\begin{lemma}[Strong Bisimulation among $ \mathcal{S}(\cdot) $]
	\label{thm:CEKS-bisim}
	If 
	$ C_1 $ and $ C_2 $ are instances of Lazy D Cast ADT,
	$ s_1, s_1' \in S(C_1) $ and $ s_2,s_2' \in S(C_2) $,
	$ s_1 \approx s_2 $,
	$ \judgeSreduce{C_1}{s_1}{s_1'} $, and
	$ \judgeSreduce{C_2}{s_2}{s_2'} $, then 
	$ s_1' \approx s_2' $
